Machine Learning guided Design\\ of Ribosome Binding Sites

### 2021-06-15 paper/main.tex
Updates from Overleaf

\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage[maxcitenames=1,style=numeric]{biblatex}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{tcolorbox}
\usepackage{graphicx}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{hyperref}
\usepackage{cleveref}
\usepackage{thmtools}   
\usepackage{thm-restate}
\usepackage{enumerate}
\usepackage{xcolor}
\usepackage{textgreek}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{authblk}
% \usepackage[table,xcdraw]{xcolor}
% \usepackage{subfloat}

@@

\title{Machine Learning guided engineering\\
of Ribosome Binding Sites}

@@

Fine control of gene expression can be achieved through engineering transcriptional and translation control elements, including the Ribosome Binding Site (RBS).
Unfortunately, RBSs are not understood at the level of finesse required for reliable design. 
To address this problem, we have created a machine learning (ML) guided engineering approach for the design of bacterial RBSs.
We used Gaussian Process Regression for prediction and the Upper Confidence Bound-based Bandit algorithm for recommendation of genetic designs to be tested in vitro.
We have integrated these ML algorithms with laboratory automation and high-throughput processes for more reliable data generation.
Our approach has shown that the algorithms were able to identify rules conducive to high protein expression.
Notably, a high number of designed RBSs demonstrate translation initiation rates equalling or exceeding the currently known strong RBSs by up to 34\%.
Overall, our results show that machine learning is a very powerful tool in design of RBSs and pave the way for engineering of more complicated genetic devices.

@@

One of the main tenets of synthetic biology is the design, evaluation and standardisation of genetic parts \cite{Brophy2014,Canton2008,Stanton2014}.
This is usually done in terms of the Design-Build-Test-Learn (DBTL) cycle, where the given genetic part or organism are continually improved by going through a number of turns of the said cycle.
This normally involves designing the DNA sequence in Computer Aided Design (CAD) software and then physically testing it in a laboratory. 
Additionally, computer modelling and prediction of part behaviour based on the designed DNA sequence or design of DNA sequence based on expected function can be used \cite{Yeoh2019,Nielsen2016}.
Most of these models are based on either the thermodynamic properties of the involved molecules (DNA, RNA, proteins, etc.) or empirically obtained values describing a relevant to a given design property, like Translation Initiation Rate (TIR) in the case of Ribosome Binding Sites (RBS) \cite{Xia1998,Chen2013,Reeve2014}.
However, de-novo design of small genetic elements is still challenging due to unknown relationships between sequence and performance of such elements.\\

The biggest gap in the DTBL cycle, at present, is at the Learn and Design interface - it is hard to translate obtained results into new designs.
For example, according to  \textcite{Reeve2014} there are three main RBS calculators, all predicting the TIR based on the thermodynamic properties of the RBS and the ribosome \cite{Seo2013,Na2010,Salis2009}. 
Reported predictions from all of these models are relatively good ($R^2 >0.8$), 
but they come with a number of caveats: i) they rely on calculations of free energies that can be hard to estimate with high precision ii) in general, one of the best ways to improve the models' accuracy is by increasing the number of phenomenons taken into account, but this can lead to paradoxically decreased model accuracy due to accumulation of errors \cite{EspahBorujeni2016} and iii) by using deterministic coefficients to calculate energies one disregards often stochastic nature of processes in the cells which again increases perceived prediction error \cite{Goss1998}. 
There are also sources showing that binding energy calculations may be poor predictors of RBS strength \cite{Saito2020,Sherer1980}. This is reinforced by studies suggesting that RNA secondary structure is potentially a more important feature in TIR determination \cite{DESMIT1994,EspahBorujeni2016}.\\

Synthetic biology is currently going through a phase of exponential increase in volume of data produced during experiments \cite{Freemont2019}. 
These new data sets can be combined with data reliant machine learning algorithms to generate new models and predictors for use in synthetic biology, vastly improving the DBTL cycle's performance \cite{Camacho2018,Radivojevic2020,LAWSON2021}. 
For example, Jervis \emph{et al.} used support vector machines and neural network to optimise production of monoterpenoid in \emph{Esherichia coli} \cite{Jervis2019}.
Similarly, Costello \emph{et al.} have used a number of machine learning approaches to analyse time-series multiomics data to predict metabolic pathway behaviour \cite{Costello2018}.
There were also successful attempts at using deep learning techniques for analysis of big synthetic biology data sets \cite{Alipanahi2015,Angermueller2016}.\\

In this work, we present how machine learning algorithms can be used as part of the DBTL cycle to predict (Learn) and recommend (Design) variants of bacterial RBS with the goal of optimising associated protein expression level. 
RBS being one of the key genetic elements controlling protein expression and at the same time having a relatively short sequence is a perfect target for establishing workflows that can be later translated to more complicated systems.
In this work we have used Gaussian Process Regression \cite{Rasmussen2004} and Upper Confidence Bound multi-armed Bandits algorithms \cite{desautels2014parallelizing} for prediction and recommendation respectively to analyse and optimise the initiation rates of the designed RBS .
Our overall experimental goal was to maximise the Translation Initiation Rate (TIR) by building and testing batches of RBS sequences while minimising the number of DBTL cycle turns that we had to do.
This way, we were able to build an extensive, reliable library of novel RBSs with diverse sequences.
At the same time we were able to discover new RBS sequences with very high TIRs from between 95 to 135\% of TIR of our chosen very strong benchmark RBS.

@@

We present our RBS optimising DBTL workflow that uses machine learning in Section~\ref{sec:dbtl-workflow}.
Machine learning (ML) is used in two different ways: we show the efficacy of the ML recommendations
in the DESIGN stage (Section~\ref{sec:ucb-results}),
and we demonstrate that the ML predictions are accurate in the LEARN stage (Section~\ref{sec:gp-results}).
We present our new RBS sequence library in Section~\ref{sec:characteristics-of-library} and describe some interesting
characteristics of the discovered sequences, 
as well as show the effectiveness of the automated laboratory workflow stages.

@@

We show our DBTL workflow, that uses machine learning to optimise protein expression, in Figure \ref{fig: Flowchart}.
BUILD and TEST are driven chiefly by choices made by human researchers and use of automated methods.
Machine learning algorithms are applied in LEARN and DESIGN.
In LEARN phase, we use the Gaussian Process regression algorithm to predict the TIR of RBS sequences comprising the experimental space.
In DESIGN phase, we use Upper Confidence Bound multi-armed Bandit algorithm to recommend new RBS designs based on the predictions from LEARN.\\

\begin{figure}[h]
    \centering
    \includegraphics[scale=0.7]{plots/Main_Paper/flowchart.pdf}
    \caption{\textbf{Flowchart of the machine learning based experimental design.} The RBS design is recommended by the Upper Confidence Bound Bandit algorithm. After generating the recommendations the RBS are built and tested using automated laboratory methods allowing for rapid construction and testing at scale. Finally, the obtained results are fed back to the Gaussian Process Regression prediction algorithm in the learn phase. }
    \label{fig: Flowchart}
\end{figure}

In our genetic design, the investigated RBS controls expression of the Green Fluorescent Protein (GFP). 
By controlling expression of a fluorescent protein with the RBS we can quickly assess the perceived relative TIR by measuring fluorescence of cells harbouring plasmid with the device over time.
Finally, the mRNA is transcribed from an IPTG-inducible promoter pLlacO-1. 
By making the whole device inducible we can synchronise the start of the expression of the GFP in all the cultures by inducing them at the same time with addition of IPTG.\\

Our template RBS sequence is 20 bps long with the sequence TTTAAGA\textbf{AGGAGA}TATACA, where the highlighted nucleotides constitute the core of the RBS.
This sequence is known to have a very high TIR and comes with the pBb series plasmids \cite{Lee2011}. 
Since this is the sequence against which new RBS sequences will be benchmarked,
we will refer to this sequence as the \textit{benchmark sequence} from now on.
We have experimentally confirmed that modifying the core sequence is statistically more impactful on TIR than changes made outside of it (see Supplementary materials).
In effect, in our design we focus on modification of the core at positions -8 to -13 (relative to the start codon of the GFP) nucleotides of the RBS and fix others to be the same as the benchmark sequence, i.e. TTTAAGA + NNNNNN + TATACAT, where N can be any nucleotide (A, C, G, T). 
The total experimental (variant) space is then $4^6$ = 4096.

@@

The design recommendations were made using the Multi-armed Bandit algorithm.
In short, this algorithm is a stochastic method of probing of the experimental space. 
This algorithm aims at maximising the reward (output) from testing a limited number of instances from a big pool which cannot be wholly tested due to limited resources (time, computational power, capital).
It also balances the exploration-exploitation paradigm, where exploration focuses on testing data points which maximise information gain and exploitation focuses on recommending RBS with high predicted TIR.\\

Figure \ref{fig: Swarmplot and Quantplot}A shows the results for all the RBS groups tested experimentally. 
In each experimental round, apart from the new RBS designs, we measure the TIR of benchmark RBS as the internal standard. 
We then obtain the normalised TIR (called \textit{TIR ratio}) by taking the ratio between the raw TIR of a new design and the average TIR of benchmark sequences in each round (which are run in triplicate in each round).\\

\begin{figure}[!ht]
    \centering
    \begin{subfigure}[b]{0.48\textwidth}
        \centering
        \caption{}
        \includegraphics[scale=0.35]{plots/Main_Paper/swarmplot.pdf}
    \end{subfigure}
    % \hfill
    \begin{subfigure}[b]{0.25\textwidth}
        \centering
        \caption{}
        \includegraphics[scale=0.35]{plots/Main_Paper/quantplot.pdf}
    \end{subfigure}
    % \hfill
    \begin{subfigure}[b]{0.25\textwidth}
        \centering
        \caption{}
        \includegraphics[scale=0.35]{plots/Main_Paper/swarmplot_proj.pdf}
    \end{subfigure}
    \begin{subfigure}[b]{0.48\textwidth}
        \centering
        \caption{}
        \includegraphics[scale=0.4]{plots/Main_Paper/histogram.pdf}
    \end{subfigure}
    \begin{subfigure}[b]{0.48\textwidth}
        \centering
        \caption{}
        \includegraphics[scale=0.42]{plots/Main_Paper/tsneplot.pdf}
    \end{subfigure}
    % \includegraphics[scale=0.3]{plots/Main_Paper/swarmplot.pdf}
    % \includegraphics[scale=0.3]{plots/Main_Paper/quantplot.pdf}
    % \includegraphics[scale=0.3]{plots/Main_Paper/swarmplot_proj.pdf}
    % \includegraphics[scale=0.3]{plots/Main_Paper/histogram.pdf}
    % \includegraphics[scale=0.4]{plots/Main_Paper/tsneplot.pdf}
    \caption{
    \textbf{TIRs of RBS groups examined in this study.} 
    \textbf{A)} Swarm plot showing the obtained TIRs divided into RBS groups.
    BPS-NC: base-by-base changes in the non-core region. 
    BPS-C: base-by-base changes in the core region. 
    UNI: Randomly generated sequences with uniform distribution. 
    PPM: Randomly generated sequences with distribution following the PPM for all natural RBS in \emph{E. coli}. 
    Bandit-0/1/2/3 - Bandit algorithm generated results for Round 0, 1, 2 and 3 respectively.
    SD - Shine-Dalgarno sequence.
    Dash line is set to 1 and represents the averaged benchmark sequence TIR for that group. 
    BN - benchmark sequences for all plates. 
    They are not all exactly 1 due to them being shown as separate samples rather than per round averages.
    \textbf{B)} Line plot showing TIR obtained in a given quantile of results divided into groups as in A).
    % save the random groups which were shown together due to similar distributions.
    UNI and PPM are merged into Random group and BPS-NC is not shown due to changes being made outside the core in that group. 
    \textbf{C)} Exploitation v.s. Exploration for Bandit 1-3. Blue-hued points represent exploitation, those hued red represent exploration. 
    \textbf{D)} Histogram with kernel density estimations (KDE) showing distributions of TIRs for Bandit groups.
    \textbf{E)} t-SNE plot showing the relative distances between sequences in our design spaces as calculated by our kernel function (weighted degree kernel with shift). 
    The area of the circle corresponds to the the experimentally obtained TIR value.
    The TIR results in all subplots are shown normalised to the respective benchmark sequence sample which acts as internal standard, that is TIR of a given RBS is divided by TIR of the benchmark RBS run in the same plate. }
    \label{fig: Swarmplot and Quantplot}
\end{figure}

To generate the data set that the algorithm could learn from we have decided to characterise a total of 450 RBS variants, which constitutes a little over 10\% of the whole experimental space. 
To fit into our automated workflow, we have divided the 450 variants into batches of 90 split into 4 design rounds.\\

In the zeroth round we have tested two batches of designs, for total of 180 variants split as below:

\begin{itemize}
    \item BPS-NC and BPS-C group: 60 RBS sequences which are subsequent single nucleotide variations of all 20 nucleotides of the original, consensus sequence. This batch is designed to show us influence of such single nucleotide changes on the overall performance of the RBS and the potential impact of changes made beyond the core part (see Supplementary Figure \ref{fig:core_vs_noncore}).
    % \mengyan{put it in main paper?}
    \item UNI group: 30 RBS sequences that were  uniformly randomised, i.e. equal probability of choosing any nucleotide for each position. 
    \item PPM group: 30 RBS sequences randomised based on the position probability matrix (PPM) generated from all the naturally occurring RBS sequences in \emph{E. coli} genome \cite{Stormo1982}.
    \item Bandit-0: 60 RBS sequences recommended by our implementation of the recommendation algorithm based on a data set obtained from literature \cite{jervis2018machine}, which contains 113 non-repeated records for 56 unique RBS sequences with the respective TIRs.
    This data set has been used due to perceived similarity of its goal to the one of this work - prediction of TIR based on phenotypic output.
\end{itemize}

@@

All Round 0 groups (BPS-NC, BPS-C, UNI, PPM, Bandit-0) have performed worse than our benchmark sequence in terms of TIR. 
The best performing of these groups is the BPS-NC, which is explained by relatively small impact on the TIR by the changes made outside the RBS core. 
The Bandit-0 group's performance is similar to randomly generated designs, despite being machine learning driven, due to being trained on only approximate data.\\

However, starting from Round 1, where the prediction and recommendation algorithms were fed data from the Round 0 the results improve significantly, with a number of sequences that perform better than the consensus Shine-Dalgarno sequence and in one case better than the benchmark (by 8\%).
In round 2 we have observed further improvement by getting more sequences that showed TIR on levels similar to our benchmark sequence.
Finally, in round 3 the algorithm identified two sequences that were 34\% and 15\% stronger than the benchmark sequence.\\

Figure \ref{fig: Swarmplot and Quantplot}B shows the same results but divided into quantiles where the specific point for a given group is showing the highest TIR for that quantile.
The gradual increase for all quantiles can be observed for all Bandit groups suggesting algorithms' better understanding of the experimental space with more data.
The decreased result in the 0.9th qunatile compared to the max value for Bandit 3 group can be attributed to the increased emphasis on exploitation that has been set for the that round compared to others.
We see that effect in Figure \ref{fig: Swarmplot and Quantplot}C, where we coloured the data points for Bandit 1-3 groups according to their relative exploration - exploitation affinity.
Those with high predicted mean are coloured blue and represent exploitation, those hued red are with high predicted uncertainty and represent exploration.
We can see that the RBSs with high TIRs tend to come from exploitation of the design space whereas the explorative points give relatively low TIR, but expand our knowledge about the unknown part of the design space.\\

Figure \ref{fig: Swarmplot and Quantplot}D shows the TIRs of RBSs tested in the Bandit groups divided into bins of width of 0.1 TIR ratio.
KDE plots have been overlaid to depict the calculated density for each group.
The increase of prevalence of later bandit groups in the higher bins is evident, especially for Bandit 2 and 3 constituting the bulk of results in the $>0.8$ TIR ratio bins.
Notably, the distributions calculated for all the groups show bimodal distribution, we discuss the possible reasons for that further in the text.\\

In \ref{fig: Swarmplot and Quantplot}E we show a t-SNE plot depicting the experimental space.
Each RBS is located on the plot according to its distance to other RBSs as calculated by our embedding function.
We can see the RBSs recommended by Bandit groups have covered majority of the design space. 
Additionally, a number of clusters were especially targeted by our recommendation algorithm.
For example, the circled clusters labelled as ``G-Rich Clusters" have been actively recommended by the algorithm.
More specifically, sequences with more or equal to 4 guanines in any position constituted 10\% of the randomly selected sequences and 5, 9, 16 and finally 25\% in each of the 4 Bandit guided batches respectively.

@@

In this work, we have shown how machine learning and high-throughput, automated laboratory methods can be used to efficiently generate a library of small parts, in this case bacterial RBS. 
We have used Gaussian Process regression to predict the shape of our function and Upper Confidence Bound Bandit algorithm to recommend sequences to be tested.
We have investigated a number of methods of digitising the DNA sequence, finally settling on Weighted Degree Kernel with Shift method, which fit well into our prediction method.
In building and testing, we have performed bulk of our experiments using automation to increase their speed, reliability and reproducibility.
By using our workflow, we have found very strong RBS sequences and we have generated an extensive library of diverse RBS that can be used in the future studies.
We have achieved this despite the relatively low accuracy of our predictions, which means that the presented algorithms are robust and able to identify the right signal in a noisy environment. \\

@@

All code and data required to reproduce the results is available at Github: \url{https://github.com/mholowko/SynbioML} .
All the processed and raw data is included in the repository.
Sequences of plasmids and oligos and assembly reports used in this study are available in supplementary as a separate file.

>>>>

\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage[maxcitenames=1,style=numeric, sorting = none]{biblatex}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{tcolorbox}
\usepackage{graphicx}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{hyperref}
\usepackage{cleveref}
\usepackage{thmtools}   
\usepackage{thm-restate}
\usepackage{enumerate}
\usepackage{xcolor}
\usepackage{textgreek}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{authblk}
% \usepackage[table,xcdraw]{xcolor}
% \usepackage{subfloat}

@@

\title{Machine Learning guided Design\\
of Ribosome Binding Sites}

@@

% Fine control of gene expression can be achieved through engineering transcription and translation control elements, including the ribosome binding site (RBS).
Ribosome binding site (RBS), as one of the engineering transcription and translation control elements, allows fine control of gene expression.
% Unfortunately, RBSs are not understood at the level of finesse required for reliable design.
We consider the problem of designing RBSs in nucleotides level to optimise the gene expression with fixed budget. 
To address this problem, we have created a machine learning (ML) guided Design-Build-Test-Learn (DBTL) cycle for the experimental design of bacterial RBSs.
We used Gaussian Process Regression for prediction and the batch Upper Confidence Bound-based Bandit algorithm for recommendation of genetic designs to be tested in vivo.
We have integrated these ML algorithms with laboratory automation and high-throughput processes for more reliable data generation.
Our approach has shown that the algorithms were able to identify rules conducive to high protein expression.
Notably, by testing only 450 RBS variants, we have discovered 
% designed and experimentally-validated 
RBSs with high translation initiation rates equalling or exceeding the currently known strong RBSs by up to 34\%.
Overall, our results show that machine learning is a very powerful tool in design of RBSs and pave the way for the experimental design of more complicated genetic devices.

@@

One of the main tenets of synthetic biology is the design, evaluation and standardisation of genetic parts \cite{Brophy2014,Canton2008,Stanton2014}.
This is usually done in terms of the Design-Build-Test-Learn (DBTL) cycle, where the given genetic part or organism is continually improved through an iterative process.
This cycle involves designing the DNA sequence in computer-aided design (CAD) software and then physically testing it in a laboratory. 
Additionally, computer modelling can be used to predict characteristics of a genetic part \cite{Yeoh2019,Nielsen2016}.
Most of these computer models are based on either the thermodynamic properties of the involved molecules (DNA, RNA, proteins, etc.) or empirically-obtained values describing a relevant design property, like translation initiation rate (TIR) in the case of ribosome binding sites (RBS) \cite{Xia1998,Chen2013,Reeve2014}.
However, de-novo design of small genetic elements is still challenging due to unknown relationships between their sequence and performance.\\

The biggest gap in the DTBL cycle, at present, is at the interface between Learn and Design - it is hard to translate obtained results into new designs.
For example, according to  \textcite{Reeve2014} there are three main RBS calculators, all predicting the TIR based on the thermodynamic properties of the RBS and the ribosome \cite{Seo2013,Na2010,Salis2009}. 
Reported predictions from all of these models are relatively good ($R^2 >0.8$), 
but they come with a number of caveats: i) they rely on calculations of free energies which can be difficult to estimate with high precision, ii) in general, one of the best ways to improve the models' accuracy is by increasing the number of phenomena taken into account, which can lead paradoxically to decreased model accuracy due to accumulation of errors \cite{EspahBorujeni2016}, and iii) by using deterministic coefficients to calculate energies, one disregards the often stochastic nature of processes in cells which increases perceived prediction error \cite{Goss1998}. 
There is also evidence that binding energy calculations may be poor predictors of RBS strength \cite{Saito2020,Sherer1980}. This is reinforced by studies suggesting that RNA secondary structure is potentially a more important feature in TIR determination \cite{DESMIT1994,EspahBorujeni2016}.\\

Currently, the volume of experimental data in synthetic biology is increasing exponentially. \cite{Freemont2019}. 
Recent work has combined these data sets with machine learning algorithms to generate new models and predictors for use in synthetic biology, vastly improving the DBTL cycle's performance \cite{Camacho2018,Radivojevic2020,LAWSON2021}. 
For example, Jervis \emph{et al.} used support vector machines and neural network to optimise production of monoterpenoids in \emph{Esherichia coli} \cite{Jervis2019}.
Similarly, Costello \emph{et al.} have used a number of machine learning approaches to analyse time-series multiomics data to predict metabolic pathway behaviour \cite{Costello2018}. Deep learning techniques have also been successfully used to analyse large synthetic biology data sets \cite{Alipanahi2015,Angermueller2016,Hollerer2020}.\\

In this work, we demonstrate how machine learning algorithms can be incorporated into the DBTL cycle to predict (Learn) and recommend (Design) variants of a bacterial RBS with the goal of optimising associated protein expression level. 
We used Gaussian Process Regression \cite{Rasmussen2004} and Upper Confidence Bound multi-armed Bandits algorithms \cite{desautels2014parallelizing} for prediction and recommendation respectively to analyse and optimise the initiation rates of the designed RBS.
As the RBS is one of the key genetic elements controlling protein expression and has a relatively short sequence, it is a perfect target for establishing a workflow that can be later translated to more complicated systems.
Our overall experimental goal was to maximise the Translation Initiation Rate (TIR) by building and testing batches of RBS sequences while minimising the number of DBTL cycle iterations.
This way, we have built an extensive and reliable library of novel RBSs with diverse sequences. We discovered new RBS sequences with very high TIRs, between 95 to 135\% of our very strong benchmark RBS.

@@

We present our RBS-optimising DBTL workflow that uses machine learning in Section~\ref{sec:dbtl-workflow}.
Machine learning is used in two different ways: i) we show the efficacy of the ML recommendations
in the Design stage (Section~\ref{sec:ucb-results}),
and ii) we demonstrate that the ML predictions are accurate in the Learn stage (Section~\ref{sec:gp-results}).
We present our new RBS sequence library in Section~\ref{sec:characteristics-of-library} and describe some interesting
characteristics of the discovered sequences, 
as well as show the effectiveness of the automated laboratory workflow.

@@

Our DBTL workflow, which uses machine learning to optimise protein expression, is shown in Figure \ref{fig: Flowchart}.
Build and Test are driven chiefly by choices made by human researchers and the use of automated methods.
Machine learning algorithms are applied in Learn and Design.
In the Learn phase, we use the Gaussian Process regression algorithm to predict the TIR of RBS sequences comprising the experimental space.
In the Design phase, we use Upper Confidence Bound multi-armed Bandit algorithm to recommend new RBS sequences based on the predictions from Learn.\\

\begin{figure}[h]
    \centering
    \includegraphics[scale=0.7]{plots/Main_Paper/flowchart.pdf}
    \caption{\textbf{Flowchart of the machine learning-based experimental design.} The RBS design is recommended by the Upper Confidence Bound Bandit algorithm. After generating the recommendations, the RBSs are built and tested using automated laboratory methods allowing for rapid construction and testing at scale. Finally, the obtained results are fed back to the Gaussian Process Regression prediction algorithm in the Learn phase. n is the current design round and k is the maximum number of rounds allowed by time and/or money.}
    \label{fig: Flowchart}
\end{figure}

In our genetic design, the investigated RBS controls expression of green fluorescent protein (GFP). 
By controlling expression of a fluorescent protein with the RBS we can quickly assess the perceived relative TIR by measuring fluorescence of cells harbouring the expression vector over time.
Finally, the mRNA is transcribed from an IPTG-inducible promoter pLlacO-1. 
Inducible expression allows one to synchronise the start of the GFP expression in all the cultures upon addition of IPTG.\\

Our template RBS sequence is 20 base pairs (bp) long with the sequence TTTAAGA\textbf{AGGAGA}TATACA, where the highlighted nucleotides constitute the core of the RBS.
This sequence is known to have a very high TIR and is present in the pBb series of plasmids \cite{Lee2011}. 
Since this is the sequence against which new RBS sequences will be benchmarked,
we will refer to this sequence as the \textit{benchmark sequence} hereafter.
We have experimentally confirmed that modifying the core sequence is statistically more impactful on TIR than changes made outside of it (see Supplementary materials).
In our design, we focus on modification of the core at nucelotide positions -8 to -13 (relative to the start codon of the GFP) of the RBS and keep other positions the same as the benchmark sequence, i.e. TTTAAGA + NNNNNN + TATACAT, where N can be any nucleotide (A, C, G, T). 
The total experimental (variant) space is then $4^6$ = 4096.

@@

The design recommendations were made using the Multi-armed Bandit algorithm.
In short, this algorithm is a stochastic method of probing of the experimental space. 
It aims at maximising the reward (output) from testing a limited number of instances from a large pool which cannot be tested exhaustively due to limited resources (time, computational power, money).
It also balances the exploration-exploitation paradigm, where exploration focuses on testing data points which maximise information gain and exploitation focuses on recommending RBSs with high predicted TIR.\\

Figure \ref{fig: Swarmplot and Quantplot}A shows the results for all the RBS groups tested experimentally. 
In each experimental round, in addition to the new RBS designs, we measure the TIR of the benchmark RBS as the internal standard. 
We then obtain the normalised TIR (called \textit{TIR ratio}) by taking the ratio between the raw TIR of a new design and the average TIR of benchmark sequences in each round (which are run in triplicate in each round).\\

\begin{figure}[!ht]
    \centering
    \begin{subfigure}[b]{0.48\textwidth}
        \centering
        \caption{}
        \includegraphics[scale=0.35]{plots/Main_Paper/swarmplot.pdf}
    \end{subfigure}
    % \hfill
    \begin{subfigure}[b]{0.25\textwidth}
        \centering
        \caption{}
        \includegraphics[scale=0.35]{plots/Main_Paper/quantplot.pdf}
    \end{subfigure}
    % \hfill
    \begin{subfigure}[b]{0.25\textwidth}
        \centering
        \caption{}
        \includegraphics[scale=0.35]{plots/Main_Paper/swarmplot_proj.pdf}
    \end{subfigure}
    \begin{subfigure}[b]{0.48\textwidth}
        \centering
        \caption{}
        \includegraphics[scale=0.4]{plots/Main_Paper/histogram.pdf}
    \end{subfigure}
    \begin{subfigure}[b]{0.48\textwidth}
        \centering
        \caption{}
        \includegraphics[scale=0.42]{plots/Main_Paper/tsneplot.pdf}
    \end{subfigure}
    % \includegraphics[scale=0.3]{plots/Main_Paper/swarmplot.pdf}
    % \includegraphics[scale=0.3]{plots/Main_Paper/quantplot.pdf}
    % \includegraphics[scale=0.3]{plots/Main_Paper/swarmplot_proj.pdf}
    % \includegraphics[scale=0.3]{plots/Main_Paper/histogram.pdf}
    % \includegraphics[scale=0.4]{plots/Main_Paper/tsneplot.pdf}
    \caption{
    \textbf{TIRs of RBS groups examined in this study.} 
    \textbf{A)} Swarm plot showing the obtained TIRs divided into RBS groups.
    BPS-NC: base-by-base changes in the non-core region. 
    BPS-C: base-by-base changes in the core region. 
    UNI: Randomly generated sequences with uniform distribution. 
    PPM: Randomly generated sequences with distribution following the PPM for all natural RBS in \emph{E. coli}. 
    Bandit-0/1/2/3 - Bandit algorithm generated results for Round 0, 1, 2 and 3 respectively.
    SD - Shine-Dalgarno sequence.
    Dash line is set to 1 and represents the averaged benchmark sequence TIR for that group. 
    BN - benchmark sequences for all plates. 
    (not all are exactly 1 due to them being shown as separate samples rather than per round averages.)
    \textbf{B)} Line plot showing TIR obtained in a given quantile of results divided into groups as in A).
    % save the random groups which were shown together due to similar distributions.
    UNI and PPM are merged into Random group and BPS-NC is not shown due to changes being made outside the core in that group. 
    \textbf{C)} Exploitation v.s. Exploration for Bandit 1-3. Blue-hued points represent exploitation, those hued red represent exploration. 
    \textbf{D)} Histogram with kernel density estimations (KDE) showing distributions of TIRs for Bandit groups.
    \textbf{E)} t-SNE plot showing the relative distances between sequences in our design spaces as calculated by our kernel function (weighted degree kernel with shift). 
    The area of the circle corresponds to the the experimentally obtained TIR value.
    The TIR results in all subplots are shown normalised to the respective benchmark sequence sample which acts as internal standard, that is TIR of a given RBS is divided by TIR of the benchmark RBS run in the same plate. }
    \label{fig: Swarmplot and Quantplot}
\end{figure}

To generate the data set that the algorithm could learn from we have decided to characterise a total of 450 RBS variants, little over 10\% of the whole experimental space. 
To fit into our automated workflow, we have divided the 450 variants into batches of 90 split into 4 design rounds.\\

In the zeroth round we have tested two batches of designs, giving a total of 180 variants split as below:

\begin{itemize}
    \item BPS-NC and BPS-C group: 60 RBS sequences which are subsequent single nucleotide variations of all 20 nucleotides of the original, consensus sequence. This batch is designed to show us influence of such single nucleotide changes on the overall performance of the RBS and the potential impact of changes made beyond the core part (see Supplementary Figure \ref{fig:core_vs_noncore}).
    % \mengyan{put it in main paper?}
    \item UNI group: 30 RBS sequences that were  uniformly randomised, i.e. equal probability of choosing any nucleotide for each position. 
    \item PPM group: 30 RBS sequences randomised based on the position probability matrix (PPM) generated from all the naturally occurring RBS sequences in \emph{E. coli} genome \cite{barrick1994quantitative}.
    % \cite{Stormo1982}.
    \item Bandit-0: 60 RBS sequences recommended by our implementation of the recommendation algorithm based on a data set obtained from literature \cite{jervis2018machine}, which contains 113 non-repeated records for 56 unique RBS sequences with the respective TIRs.
    This data set has been used due to perceived similarity of its goal to the one of this work - prediction of TIR based on phenotypic output.
\end{itemize}

@@

All Round 0 groups (BPS-NC, BPS-C, UNI, PPM, Bandit-0) have performed worse than our benchmark sequence in terms of TIR. 
The best performing of these groups is the BPS-NC, which is explained by relatively small impact on the TIR of changes made outside the RBS core. 
The Bandit-0 group's performance is similar to randomly generated designs, despite being machine learning driven, due to being trained on approximate data.\\

However, starting from Round 1, where the prediction and recommendation algorithms were fed data from Round 0, the results improve significantly, with a number of sequences which perform better than the consensus Shine-Dalgarno sequence and in one case better than the benchmark (by 8\%).
In Round 2 we have observed further improvement by obtaining more sequences with TIRs similar to our benchmark sequence.
Finally, in Round 3 the algorithm identified two sequences that were 34\% and 15\% stronger than the benchmark sequence.\\

Figure \ref{fig: Swarmplot and Quantplot}B shows the same results but divided into quantiles where the specific point for a given group shows the highest TIR for that quantile.
The gradual increase for all quantiles can be observed for all Bandit groups, suggesting that the algorithms have a better understanding of the experimental space given more data.
The decreased result in the 0.9th quantile compared to the max value for Bandit 3 group can be attributed to the increased emphasis on exploitation that has been set for that round compared to others.
We see this effect in Figure \ref{fig: Swarmplot and Quantplot}C, where we coloured the data points for Bandit 1-3 groups according to their relative exploration - exploitation affinity.
Those with a high predicted mean are coloured blue and represent exploitation, those coloured red are with high predicted uncertainty and represent exploration.
We can see that the RBSs with high TIRs tend to come from exploitation of the design space, whereas the explorative points give relatively low TIRs, but expand our knowledge of the unknown part of the design space.\\

Figure \ref{fig: Swarmplot and Quantplot}D shows the TIRs of RBSs tested in the Bandit groups divided into bins with width equal to a TIR ratio of 0.1.
KDE plots have been overlaid to depict the calculated density for each group.
The increase in prevalence of later bandit groups in the higher bins is evident, especially for Bandit 2 and 3, constituting the bulk of results in the $>0.8$ TIR ratio bins.
Notably, the distributions calculated for all the groups are bimodal - we discuss the possible reasons for that later in the text.\\

In \ref{fig: Swarmplot and Quantplot}E we show a t-SNE plot depicting the experimental space.
Each RBS is located on the plot according to its distance from other RBSs as calculated by our embedding function.
The RBSs recommended by Bandit groups have covered majority of the design space. 
Additionally, a number of clusters were especially targeted by our recommendation algorithm.
For example, the circled clusters labelled as ``G-Rich Clusters" have been actively recommended by the algorithm.
More specifically, sequences with more or equal to 4 guanines in any position constituted 10\% of the randomly selected sequences and 5, 9, 16 and finally 25\% in each of the 4 Bandit guided batches respectively.

@@

In this work, we have shown how machine learning and high-throughput, automated laboratory methods can be jointly applied to efficiently generate a library of small parts, in this case bacterial RBS. 
We have used Gaussian Process regression to predict the TIR function to be optimised and Upper Confidence Bound Bandit algorithm to recommend sequences to be tested.
% We have investigated a number of methods of digitising the DNA sequence, finally settling on 
To represent RBSs and capture the similarities between them, we chose to use the
Weighted Degree Kernel with Shift method, which fit well into our prediction method.
In building and testing, we have performed bulk of our experiments using automation to increase their speed, reliability and reproducibility.
By using our workflow and testing only 450 RBS variants, we have discovered 
% designed and experimentally-validated 
RBSs with high translation initiation rates equalling or exceeding the currently known strong RBSs by up to 34\%.
Furthermore, we have generated an extensive library of diverse RBS that can be used in the future studies.\\

Our approach has shown that machine learning is able to correctly detect and exploit rules of biological design that otherwise require a substantial time and experiments to uncover.
For example, our algorithm has correctly identified the correlation of high guanine content in the RBS with high TIR of such sequences.
We have achieved this despite the relatively low accuracy of our predictions, which benefits from recommendation algorithms with the exploitation-exploration balance.
This gives hope that the continuing rise in use of machine learning in synthetic biology will enable to uncover new biologically relevant rules of design.\\

@@

All code and data required to reproduce the results is available at Github: \url{https://github.com/mholowko/Solaris/tree/master/synbio_rbs} .
All the processed and raw data is included in the repository.
Sequences of plasmids and oligos and assembly reports used in this study are available in supplementary as a separate file.
